Informing policy via dynamic models: Eliminating cholera in Haiti

### 2022-08-05 ms.Rnw
Replaced all stew calls with bakes (except one). Should work on other machines with required .rda files

true_agg_cases <- haitiCholera %>%
  select(-report) %>%
  pivot_longer(
    data = .,
    cols = -date_saturday,
    values_to = 'cases',
    names_to = 'dep'
  ) %>%
  mutate(
    date = as.Date(date_saturday),
    dep = gsub("\\.", "_", dep)
  ) %>%
  mutate(
    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
  ) %>%
  tidyr::pivot_wider(
    data = .,
    id_cols = c(date),
    names_from = dep,
    values_from = cases,
    names_prefix = 'cases_'
  ) %>%
  mutate(
    ReportedAll = cases_Artibonite + cases_Centre +
      cases_Grande_Anse + cases_Nippes + cases_Nord +
      cases_Nord_Est + cases_Ouest + cases_Sud +
      cases_Sud_Est + cases_Nord_Ouest
  )

@@

<<Model 3 PanelSims,echo=FALSE>>=
departements <-
  c(
    'Artibonite',
    'Centre',
    'Grande_Anse',
    'Nippes',
    'Nord',
    'Nord-Est',
    'Nord-Ouest',
    'Ouest',
    'Sud',
    'Sud-Est'
  )

@@

stew(
  file = "model3/output/sims_FinalPanel.rda", {
    foreach(dep=departements, .combine = rbind) %dopar% {
      SIRB <- unitobjects(h3)[[dep]]
      shared <- h3@shared
      specific <- h3@specific[, dep]

      pomp::simulate(SIRB,
                     params = c(shared, specific),
                     nsim = 500, format = 'data.frame',
                     include.data = TRUE) -> sims

      sims$dep <- dep
      sims
    } -> h3_panel_all_sims

    h3_panel_quantiles <- h3_panel_all_sims %>%
      mutate(isdata = .id == 'data') %>%
      # filter(!isdata) %>%
      group_by(isdata, dep, time) %>%
      summarise(
        q05 = quantile(cases, 0.025, na.rm = T),
        mean = mean(cases, na.rm = T),
        q50 = quantile(cases, 0.5, na.rm = T),
        q95 = quantile(cases, 0.975, na.rm = T)
      )

    h3_panel_real_data <- h3_panel_quantiles %>%
      filter(isdata)

    h3_panel_sim_data <- h3_panel_quantiles %>%
      filter(!isdata)

  }
)

h3_panel_sim_data <- h3_panel_sim_data %>%
  mutate(
    dep = gsub("-", "_", dep),
    date = yearsToDateTime(time)
  ) %>%
  filter(date >= yearsToDate(h3@unit.objects$Artibonite@t0)) %>%
  mutate(date = as.Date(lubridate::round_date(date)))

h3_panel_real_data <- h3_panel_real_data %>%
  mutate(
    dep = gsub("-", "_", dep),
    date = yearsToDateTime(time)
  ) %>%
  filter(date >= yearsToDate(h3@unit.objects$Artibonite@t0)) %>%
  mutate(date = as.Date(lubridate::round_date(date)))

@@

\begin{figure}[ht]
<<Model3_PanelSims_Figure, echo=FALSE, fig.height=3>>=

ggplot() +
  geom_line(data = h3_panel_sim_data, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_line(data = h3_panel_real_data, aes(x = date, y = q50 + 1), col = 'black') +
  geom_ribbon(data = h3_panel_sim_data, aes(x = date, ymin = q05 + 1, ymax = q95 + 1),
              alpha = 0.4) +
  facet_wrap(~factor(dep, levels = plot_order), nrow = 2,
             labeller = dep_labeller) +
  ylab('Reported cholera cases') +
  theme(axis.title.x = element_blank()) +
  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2011-01-01"), as.Date("2019-01-01"), by = '2 years')) +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
  # theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1,
  #                                  size = 7),
        # axis.title.x = element_blank())
@
\caption{\label{h3panelsims}
Simulations from initial conditions using panelPOMP version of Model~3.
The black curve represents true case count, the blue curve the median of 500 simulations from the model, and the grey ribbons representing $95\%$ confidence interval.
}
\end{figure}

<<Remove Model3 panelSims, cache=FALSE, include=FALSE, message=FALSE, echo=FALSE>>=
rm(h3_panel_all_sims, h3_panel_sim_data)
gc()
@

@@

h3Spat <- haiti3_spatPomp()
coef(h3Spat) <- haitipkg:::panelParms_toSpatParms3(p3)

registerDoRNG(6687147)

stew(
  file = "model3/output/SpatPompSims.rda", {
    h3Spat_sims <- simulate(h3Spat, nsim = 500, seed = 784713,
                     format = "data.frame", include.data = TRUE)

    h3Spat_quants <- h3Spat_sims %>%
      mutate(is.data = .id == 'data') %>%
      filter(!is.data) %>%
      rename(dep = unitname) %>%
      group_by(dep, time) %>%
      summarise(q05 = quantile(cases, 0.025, na.rm = T),
                mean = mean(cases, na.rm = T),
                q50 = quantile(cases, 0.5, na.rm = T),
                q95 = quantile(cases, 0.975, na.rm = T)) %>%
      ungroup() %>%
      mutate(date = yearsToDateTime(time)) %>%
      filter(date >= yearsToDate(h3Spat@t0)) %>%
      mutate(date = as.Date(lubridate::round_date(date)))

    h3Spat_truth <- h3Spat_sims %>%
      # select(time, .id, starts_with("cases")) %>%
      mutate(is.data = .id == 'data') %>%
      filter(is.data) %>%
      rename(dep = unitname) %>%
      group_by(dep, time) %>%
      summarise(q05 = quantile(cases, 0.025, na.rm = T),
                mean = mean(cases, na.rm = T),
                q50 = quantile(cases, 0.5, na.rm = T),
                q95 = quantile(cases, 0.975, na.rm = T)) %>%
      ungroup() %>%
      mutate(date = yearsToDateTime(time)) %>%
      filter(date >= yearsToDate(h3Spat@t0)) %>%
      mutate(date = as.Date(lubridate::round_date(date)))

  }
)

@@

\begin{figure}[ht]
<<Model3_SpatPOMP_sims_Figure, fig.height=3>>=
ggplot() +
  geom_line(data = h3Spat_quants, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_line(data = h3Spat_truth, aes(x = date, y = q50 + 1), col = 'black') +
  geom_ribbon(data = h3Spat_quants,
              aes(x = date, ymin = q05 + 1, ymax = q95 + 1),
              alpha = 0.4) +
  facet_wrap(~factor(dep, levels = plot_order), nrow = 2,
             labeller = dep_labeller) +
  labs(y = 'Log Reported Cholera Cases') +
  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2011-01-01"), as.Date("2019-01-01"), by = '2 years')) +
  theme(axis.title.x = element_blank()) +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
  # theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1,
  #                                  size = 6),
  #       axis.title.x = element_blank())
@
\caption{\label{h3spatsims}
Simulations from initial conditions using SpatPOMP version of Model~3.
The black curve represents true case count, the blue line the median of 500 simulations from the model, and the grey ribbons representing $95\%$ confidence interval.
}
\end{figure}

<<Remove Model3 spatSims, cache=FALSE, include=FALSE, message=FALSE, echo=FALSE>>=
rm(h3Spat_sims, h3Spat_truth, h3Spat_quants)
gc()
@

@@

<<Lee3 Sub Calculations, echo=FALSE, include=FALSE, message=FALSE>>=
registerDoRNG(49637421)

@@

stew(
  # Save the file
  file = 'model3/output/lee3_results.rda', {

@@

      # Save all likelihood evaluations
      lee3_results[["sub_ll_evals"]] <- foreach(
        i = 1:36, .combine = c
      ) %dopar% {
        logLik(pfilter(lee3_sub, Np = 5000))  # Get likelihood evaluations
      }
    })

@@

      # Save all likelihood evaluations
      lee3_results[["all_ll_evals"]] <- foreach(
        i = 1:36, .combine = c
      ) %dopar% {
        logLik(pfilter(lee3_all, Np = 5000))  # Get likelihood evaluations
      }
    })

    lee3_results[["n_params"]] <- 29
  }
)

@@

# Clean-up simulations for analysis/visualization
mod1_V4_sims <- agg_mod1_sims(h1_S1_sims)
mod1_V4_probs <- get_elimProbs(h1_S1_sims, model = 1)
rm(h1_S1_sims)

>>>>

dep_plot_df <- haitiCholera %>%
  select(-report) %>%
  pivot_longer(
    data = .,
    cols = -date_saturday,
    values_to = 'cases',
    names_to = 'dep'
  ) %>%
  mutate(
    date = as.Date(date_saturday),
    dep = gsub("\\.", "_", dep)
  ) %>%
  mutate(
    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
  )

true_agg_cases <- dep_plot_df %>%
  tidyr::pivot_wider(
    data = .,
    id_cols = c(date),
    names_from = dep,
    values_from = cases,
    names_prefix = 'cases_'
  ) %>%
  mutate(
    ReportedAll = cases_Artibonite + cases_Centre +
      cases_Grande_Anse + cases_Nippes + cases_Nord +
      cases_Nord_Est + cases_Ouest + cases_Sud +
      cases_Sud_Est + cases_Nord_Ouest
  )

@@

<<Model 3 PanelSims,echo=FALSE>>=
h3_panel_sims <- switch(RUN_LEVEL, 20, 100, 500)

departements <-
  c(
    'Artibonite',
    'Centre',
    'Grande_Anse',
    'Nippes',
    'Nord',
    'Nord-Est',
    'Nord-Ouest',
    'Ouest',
    'Sud',
    'Sud-Est'
  )

@@

h3_panel_sims <- bake(
  file = paste0('model3/', rl_dir, 'panelPompSims.rds'), {
    foreach(dep = departements, .combine = rbind) %dopar% {
      SIRB <- unitobjects(h3)[[dep]]
      shared <- h3@shared
      specific <- h3@specific[, dep]

      pomp::simulate(
        SIRB,
        params = c(shared, specific),
        nsim = h3_panel_sims,
        format = 'data.frame'
      ) -> sims

      sims$dep <- dep
      sims
    }
  }
)

h3_panel_quantiles <- h3_panel_sims %>%
  mutate(
    date = as.Date(
      lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
    )
  ) %>%
  group_by(dep, date) %>%
  summarise(
    q05 = quantile(cases, 0.025, na.rm = T),
    mean = mean(cases, na.rm = T),
    q50 = quantile(cases, 0.5, na.rm = T),
    q95 = quantile(cases, 0.975, na.rm = T)
  ) %>%
  ungroup() %>%
  mutate(
    dep = gsub("-", "_", dep)
  ) %>%
  filter(date >= yearsToDate(h3@unit.objects$Artibonite@t0))

@@

\begin{figure}[ht]
<<Model3_PanelSims_Figure, echo=FALSE, fig.height=3>>=
ggplot() +
  geom_line(data = h3_panel_quantiles, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_line(data = dep_plot_df, aes(x = date, y = cases + 1), col = 'black') +
  geom_ribbon(data = h3_panel_quantiles, aes(x = date, ymin = q05 + 1, ymax = q95 + 1),
              alpha = 0.4) +
  facet_wrap(~factor(dep, levels = plot_order), nrow = 2,
             labeller = dep_labeller) +
  ylab('Reported cholera cases') +
  theme(axis.title.x = element_blank()) +
  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2011-01-01"), as.Date("2019-01-01"), by = '2 years')) +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))

@
\caption{\label{h3panelsims}
Simulations from initial conditions using panelPOMP version of Model~3.
The black curve represents true case count, the blue curve the median of 500 simulations from the model, and the grey ribbons representing $95\%$ confidence interval.
}
\end{figure}

<<Remove Model3 panelSims, cache=FALSE, include=FALSE, message=FALSE, echo=FALSE>>=
rm(h3_panel_quantiles, h3_panel_sims)
gc()
@

@@

h3_spat_nsim <- switch(RUN_LEVEL, 20, 100, 500)

h3Spat_quants <- bake(
  file = paste0('model3/', rl_dir, 'SpatPompSims.rds'), {
    h3Spat <- haiti3_spatPomp()
    coef(h3Spat) <- haitipkg:::panelParms_toSpatParms3(p3)

    h3Spat_sims <- simulate(
      h3Spat, nsim = h3_spat_nsim, format = 'data.frame'
    )

    h3Spat_sims %>%
      rename(dep = unitname) %>%
      group_by(dep, time) %>%
      summarise(
        q05 = quantile(cases, 0.025, na.rm = T),
        mean = mean(cases, na.rm = T),
        q50 = quantile(cases, 0.5, na.rm = T),
        q95 = quantile(cases, 0.975, na.rm = T)
      ) %>%
      ungroup() %>%
      mutate(
        date = lubridate::date_decimal(time)
      ) %>%
      filter(date >= yearsToDate(h3Spat@t0)) %>%
      mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
  },
  timing = FALSE
)

@@

\begin{figure}[ht]
<<Model3_SpatPOMP_sims_Figure, fig.height=3>>=
ggplot() +
  geom_line(data = h3Spat_quants, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_line(data = dep_plot_df, aes(x = date, y = cases + 1), col = 'black') +
  geom_ribbon(data = h3Spat_quants,
              aes(x = date, ymin = q05 + 1, ymax = q95 + 1),
              alpha = 0.4) +
  facet_wrap(~factor(dep, levels = plot_order), nrow = 2,
             labeller = dep_labeller) +
  labs(y = 'Log Reported Cholera Cases') +
  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2011-01-01"), as.Date("2019-01-01"), by = '2 years')) +
  theme(axis.title.x = element_blank()) +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
@
\caption{\label{h3spatsims}
Simulations from initial conditions using SpatPOMP version of Model~3.
The black curve represents true case count, the blue line the median of 500 simulations from the model, and the grey ribbons representing $95\%$ confidence interval.
}
\end{figure}

<<Remove Model3 spatSims, cache=FALSE, include=FALSE, message=FALSE, echo=FALSE>>=
rm(h3Spat_quants, h3_spat_nsim)
gc()
@

@@

<<Lee3 Sub Calculations, echo=FALSE, include=FALSE, message=FALSE>>=
lee3_NP   <- switch(RUN_LEVEL, 50, 250, 5000)
lee3_NREP <- switch(RUN_LEVEL,  3,   8,   36)

registerDoRNG(49637421)

@@

lee3_results <- bake(
  file = paste0('model3/', rl_dir, 'lee3_results.rds'), {

@@

      # Save all likelihood evaluations
      lee3_results[["sub_ll_evals"]] <- foreach(
        i = 1:lee3_NREP, .combine = c
      ) %dopar% {
        logLik(pfilter(lee3_sub, Np = lee3_NP))  # Get likelihood evaluations
      }
    })

@@

      # Save all likelihood evaluations
      lee3_results[["all_ll_evals"]] <- foreach(
        i = 1:lee3_NREP, .combine = c
      ) %dopar% {
        logLik(pfilter(lee3_all, Np = lee3_NP))  # Get likelihood evaluations
      }
    })

    lee3_results[["n_params"]] <- 29

    lee3_results
  },
  timing = FALSE
)

@@

# Clean-up simulations for analysis/visualization
mod1_V4_sims <- agg_mod1_sims(h1_S1_sims)
mod1_V4_probs <- get_elimProbs(h1_S1_sims, model = 1)
rm(h1_S1_sims, h1_scen_NP, h1_scen_sims)
